opponent modelling

### 2019-11-12 comp_model_1019.Rmd
all results added

```{r}
dat <- read.csv("data20180719.csv")
# dat <- read.csv("test_data.csv")
```

@@

```{r}
naive_bayes <- function(par,data,opp_strategy_vec,return_value=c("-2loglik","likelihood_by_trial")) {
  # par = parameter, noise in computer choice
  # dat = data subset for one participant
  # opp_strategy_vec = c("level0","level1","level2")
  # return = -2logLik 
  # or "likelihood_by_trial" vector  (for later plotting, etc)

  return_value <- match.arg(return_value)
  theta <- par[1]
  eps <- par[2]

@@

```{r}
# 
# naive_bayes_modelling <- list()
# for(id in unique(dat$human_id)) {
#   naive_bayes_modelling[[id]] <- list()
#   tdat <- subset(dat,human_id == id)
#   naive_bayes_modelling[[id]] <- optim(c(0.1,0.1),fn=naive_bayes,gr = NULL, data=tdat,opp_strategy_vec = c("level0","level1","level2"),"-2loglik", lower = c(0.01,0.01), upper = c(0.99,0.99), method="L-BFGS-B")
# }
# 
# save(naive_bayes_modelling,file="naive_bayes_modelling.RData")
# 
# ```
# ```{r}
# Nbayes_par_results <- data.frame()
# for(id in unique(dat$human_id)) {
# Nbayes_par_results <- rbind(Nbayes_par_results,
#                        data.frame(
#                          "id" = id,
#                          "condition" = dat[dat$human_id==id,"condition"][1],
#                          # Theta is the parameter governing AI stochasticity. Truth is 0.9
#                          "theta" = naive_bayes_modelling[[id]]$par[1],
#                          # Epsilone is parameter showing cases where human deviates from predictions about opponent (lower better)
#                          "eps" = naive_bayes_modelling[[id]]$par[2]))
# }
# 
# # Plotting the theta/eps scatter for all participants...
# plot(Nbayes_par_results$theta,Nbayes_par_results$eps,xlab="Theta", ylab= "Epsilone",col = ifelse(as.character(Nbayes_par_results$condition) == "Level2",'red','green'))
# 
# write.csv(Nbayes_par_results,file="Nbayes_par_results.csv",row.names = FALSE)
#kable(Nbayes_par_results)
```

## Q-learning
```{r}

Q_learn <- function(par,data,return_value=c("-2loglik","likelihood_by_trial")){
  # actions
  # beta= inverse temperature parameter in softmax choice function
  # lambda = learning rate (one per game?)
  # data : participant and ai choice data.
  # Returns Q-values per trial and predicts choice using softmax
  beta <- par[1]
  lambda <- par[2]
  Q_vals_RPS = matrix(0.0,3)
  names(Q_vals_RPS) <- c("R","P","S")
  Q_vals_FWG = matrix(0.0,3)
  names(Q_vals_FWG) <- c("F","W","G")
  Q_vals_NUM = matrix(0.0,5)
  names(Q_vals_NUM) <- c("1","2","3","4","5")

@@

```{r}
# QL_modelling <- list()
# for(id in unique(dat$human_id)) {
#   QL_modelling[[id]] <- list()
#   tdat <- subset(dat,human_id == id)
#   QL_modelling[[id]] <- optim(c(0.1,0.1),fn=Q_learn,gr = NULL, data=tdat,"-2loglik", lower = c(0.001,0.01), upper = c(0.99,10), method="L-BFGS-B")
# }
# 
# save(QL_modelling,file="QL_modelling.RData")
# 
#

@@

## Parametric Experience Weighted Attraction model (Camerer & Ho 1997)
```{r}

@@

# 
# 
EWA_par <- function(par,data,return_value=c("-2loglik","likelihood_by_trial")){
  # par: vector of parameters to the EWA model
  # Data :choice data by trial for each participant
  # Returns sum of loglikelihoods or likelihood per trial
  phi <- par[1]
  delta <- par[2]
  rho <- par[3]
  lambda <- par[4]

  # Initiate N(0) = 1 as in Camerer and Ho 1997 paper.
  N <- 1

@@

  for(t in 1:nrow(data)) {
    t_game <- data[t,"game"]
    if(t_game == "numbers") nopts <- 5 else nopts <- 3
    if(data[t,"round"] == 1) {
      # first round is uniform prediction
      Att <- switch(as.character(t_game),rps=A_RPS,fwg = A_FWG, numbers = A_NUM)
      reward <- switch(as.character(t_game),rps=reward_RPS,fwg = reward_FWG, numbers = reward_NUM)
      lik_hum[t] <- 1/nopts
    } else {
      # Get reward and past human action
      h_act_prev <- as.character(data[t-1,"h_action"])
      a_act_prev <- as.character(data[t-1,"a_action"])
      for (i in 1:length(Att)) {
        action <- as.character(names(Att)[i])
        # cat("this is current strat",action,"\n")
        # cat("this is previous human action",h_act_prev,"\n")

@@

 if(return_value == "-2loglik") {
    ret <- -2*sum(log(lik_hum))
    if(is.infinite(ret)) {
      return(1e+300)
    } else {
      return(ret)
    }
  }
  if(return_value == "likelihood_by_trial") return(lik_hum)
}
```

EWA_modelling <- list()
for(id in unique(dat$human_id)) {
  EWA_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
  # EWA_modelling[[id]] <- optim(c(0.5,0.5,0.5,0.5),fn=EWA_par,gr = NULL, data=tdat,"-2loglik", lower = c(0.001,0.001,0.001,0.001), upper = c(1,0.99,0.99,1), method="L-BFGS-B")
  EWA_modelling[[id]] <- DEoptim(fn=EWA_par,lower = c(0.001,0.001,0.001,0.001), upper = c(1,0.99,0.99,1), data=tdat,"-2loglik")
}

@@

##  Self-Tuning EWA (Camerer & Ho 2007) 
```{r}
#######################################################################################

EWA_self <- function(par,data,return_value=c("-2loglik","likelihood_by_trial")){

  lambda <- par[1]
  # Initiate N(0) = 1 as in Camerer and Ho 1997 paper. 
  N <- 1.0

@@

  for(t in 1:nrow(data)) {
    t_game <- data[t,"game"]
    if(t_game == "numbers") nopts <- 5 else nopts <- 3
    if(data[t,"round"] == 1) {
      # first round is uniform prediction
      Att <- switch(as.character(t_game),rps=A_RPS,fwg = A_FWG, numbers = A_NUM)
      reward <- switch(as.character(t_game),rps=reward_RPS,fwg = reward_FWG, numbers = reward_NUM)
      game_data <- subset(data, game == t_game)
      lik_hum[t] <- 1/nopts

@@

EWA_self_modelling <- list()
for(id in unique(dat$human_id)) {
  EWA_self_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
#   EWA_self_modelling[[id]] <- optim(10.0,fn=EWA_self,gr = NULL, data=tdat,"-2loglik", lower = 0.0, upper = 200, method="L-BFGS-B")
# }
  EWA_self_modelling[[id]] <- DEoptim(fn=EWA_self, lower = 0.0, upper = 10.0, data=tdat)
}

@@

>>>>

```{r}
dat <- read.csv("data20180719.csv")
#dat <- read.csv("test_data.csv")
```

@@

```{r}
naive_bayes <- function(par,data,opp_strategy_vec,return_value) {
  # par = parameter, noise in computer choice
  # dat = data subset for one participant
  # opp_strategy_vec = c("level0","level1","level2")
  # return = -2logLik 
  # or "likelihood_by_trial" vector  (for later plotting, etc)

  #return_value <- match.arg(return_value)
  theta <- par[1]
  eps <- par[2]

@@

naive_bayes_modelling <- list()
for(id in unique(dat$human_id)) {
  naive_bayes_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
  naive_bayes_modelling[[id]] <- optim(c(0.1,0.1),fn=naive_bayes,gr = NULL, data=tdat,opp_strategy_vec = c("level0","level1","level2"),return_value = "-2loglik", lower = c(0.01,0.01), upper = c(0.99,0.99), method="L-BFGS-B")
}

save(naive_bayes_modelling,file="naive_bayes_modelling.RData")

# ```

# ```{r}
# Nbayes_par_results <- data.frame()
# for(id in unique(dat$human_id)) {
# Nbayes_par_results <- rbind(Nbayes_par_results,
#                        data.frame(
#                          "id" = id,
#                          "condition" = dat[dat$human_id==id,"condition"][1],
#                          # Theta is the parameter governing AI stochasticity. Truth is 0.9
#                          "theta" = naive_bayes_modelling[[id]]$par[1],
#                          # Epsilone is parameter showing cases where human deviates from predictions about opponent (lower better)
#                          "eps" = naive_bayes_modelling[[id]]$par[2]))
# }
# 
# # Plotting the theta/eps scatter for all participants...
# plot(Nbayes_par_results$theta,Nbayes_par_results$eps,xlab="Theta", ylab= "Epsilone",col = ifelse(as.character(Nbayes_par_results$condition) == "Level2",'red','green'))
# 
# write.csv(Nbayes_par_results,file="Nbayes_par_results.csv",row.names = FALSE)
#kable(Nbayes_par_results)
```

## Q-learning
```{r}
Q_learn <- function(par,data,return_value=c("-2loglik","likelihood_by_trial")){
  # actions
  # beta= inverse temperature parameter in softmax choice function
  # lambda = learning rate (one per game?)
  # data : participant and ai choice data.
  # Returns Q-values per trial and predicts choice using softmax
  beta <- par[1]
  lambda <- par[2]
  Q_vals_RPS = matrix(0.0,3)
  names(Q_vals_RPS) <- c("R","P","S")
  Q_vals_FWG = matrix(0.0,3)
  names(Q_vals_FWG) <- c("F","W","G")
  Q_vals_NUM = matrix(0.0,5)
  names(Q_vals_NUM) <- c("1","2","3","4","5")

@@

##Q-learning fit
```{r}
QL_modelling <- list()
for(id in unique(dat$human_id)) {
  QL_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
  QL_modelling[[id]] <- optim(c(1,0.1),fn=Q_learn,gr = NULL, data=tdat,"-2loglik", lower = c(0.001,0.001), upper = c(10,0.99), method="L-BFGS-B")
}

save(QL_modelling,file="QL_modelling.RData")

@@

## Parametric Experience Weighted Attraction model (Camerer & Ho 1997)
```{r}
EWA_par <- function(par,data,return_value){
  # par: vector of parameters to the EWA model
  # Data :choice data by trial for each participant
  # Returns sum of loglikelihoods or likelihood per trial
  phi <- par[1]
  delta <- par[2]
  rho <- par[3]
  lambda <- par[4]

@@

@@

  for(t in 1:nrow(data)) {
    t_game <- data[t,"game"]
    if(t_game == "numbers") nopts <- 5 else nopts <- 3
    if(data[t,"round"] == 1) {
      # first round is uniform prediction
      Att <- switch(as.character(t_game),rps=A_RPS,fwg = A_FWG, numbers = A_NUM)
      reward <- switch(as.character(t_game),rps=reward_RPS,fwg = reward_FWG, numbers = reward_NUM)
      lik_hum[t] <- 1/nopts
      # Initiate N(0) = 1 as in Camerer and Ho 1997 paper.
      N <- 1

    } else {
      # Get reward and past human action
      h_act_prev <- as.character(data[t-1,"h_action"])
      a_act_prev <- as.character(data[t-1,"a_action"])
      for (i in 1:length(Att)) {
        action <- as.character(names(Att)[i])
        # cat("this is current strat",action,"\n")
        # cat("this is previous human action",h_act_prev,"\n")

@@

 if(return_value == "-2loglik") {
    ret <- -2*sum(log(lik_hum))
    if(is.infinite(ret) || is.nan(ret) ) {
      return(1e+300)
    } else {
      return(ret)
    }
  }
  if(return_value == "likelihood_by_trial") return(lik_hum)
}
```

## Fitting Parametric EWA to data 
```{r}

EWA_modelling <- list()
for(id in unique(dat$human_id)) {
  EWA_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
  # EWA_modelling[[id]] <- optim(c(0.5,0.5,0.5,0.5),fn=EWA_par,gr = NULL, data=tdat,"-2loglik", lower = c(0.001,0.001,0.001,0.001), upper = c(1,0.99,0.99,1), method="L-BFGS-B")
  EWA_modelling[[id]] <- DEoptim(fn=EWA_par,lower = c(0.01,0.01,0.01,0.01), upper = c(10,1,1,10), data=tdat,"-2loglik",control=list(trace = FALSE,parallelType=1))
}

@@

##  Self-Tuning EWA (Camerer & Ho 2007) 
```{r}

EWA_self <- function(par,data,return_value){

  lambda <- par[1]
  # Initiate N(0) = 1 as in Camerer and Ho 1997 paper.

@@

  for(t in 1:nrow(data)) {
    t_game <- data[t,"game"]
    if(t_game == "numbers") nopts <- 5 else nopts <- 3
    if(data[t,"round"] == 1) {
      # first round is uniform prediction
      Att <- switch(as.character(t_game),rps=A_RPS,fwg = A_FWG, numbers = A_NUM)
      reward <- switch(as.character(t_game),rps=reward_RPS,fwg = reward_FWG, numbers = reward_NUM)
      game_data <- subset(data, game == t_game)
      lik_hum[t] <- 1/nopts
      N <- 1.0

@@

```{r}
##Plotting self_tuning EWA likelihood function for lambda from 0 to 100

lambdas <- seq(0,100,1)
data_test <- subset(data,human_id =="oXdDF_U-Dft2ja2lAAAs")

plot(lambdas,EWA_self(lambdas,data_test,"-2loglik"))
```

## Fitting Self-Tuning EWA to data 
```{r}

EWA_self_modelling <- list()
for(id in unique(dat$human_id)) {
  EWA_self_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
#   EWA_self_modelling[[id]] <- optim(10.0,fn=EWA_self,gr = NULL, data=tdat,"-2loglik", lower = 0.0, upper = 200, method="L-BFGS-B")
# }
  EWA_self_modelling[[id]] <- DEoptim(fn=EWA_self, lower = 0.0, upper = 100.0, data=tdat, "-2loglik", control=list(trace = FALSE,parallelType=1))
}

@@

```
```{r}
cat(EWA_self_modelling[["38VxtUSv_h6RR5-tAAA2"]]$optim$bestmem)
```

## Putting results together 
```{r}
All_results <- data.frame()
for(id in unique(dat$human_id)) {
All_results <- rbind(All_results,
                       data.frame(
                         "id" = id,
                         "condition" = dat[dat$human_id==id,"condition"][1],
                         "random_BIC" = -2*(100*log(1/3) + 50*log(1/5)),

                         # Naive Bayes, Theta is the parameter governing AI stochasticity. Truth is 0.9
                         "NB_theta" = naive_bayes_modelling[[id]]$par[1],
                         # Epsilone is parameter showing cases where human deviates from predictions about opponent (lower better)
                         "NB_eps" = naive_bayes_modelling[[id]]$par[2],
                         "NB_BIC" = naive_bayes_modelling[[id]]$value + 2*log(150),

                         # Q-Learning 
                         # beta ->  inverse temperature parameter in softmax choice function
                         "QL_Beta" = QL_modelling[[id]]$par[1],
                         # lambda -> learning rate in QL update
                         "QL_beta" = QL_modelling[[id]]$par[2],
                         "QL_BIC" = QL_modelling[[id]]$value + 2*log(150),

                         # Parametric EWA
                         # Phi is depreciation of past attractions
                         "EWA_Phi" = EWA_modelling[[id]]$optim$bestmem[1],
                         # Delta is weight of foregone payoffs vs actual payoffs
                         "EWA_Delta" = EWA_modelling[[id]]$optim$bestmem[2],
                         # Rho is depreciation of the experience measure N(t)
                         "EWA_Rho" = EWA_modelling[[id]]$optim$bestmem[3],
                         #Lambda is a parameter of the softmax choice function (inverse Temperature)
                         "EWA_Lambda" = EWA_modelling[[id]]$optim$bestmem[4],
                         # Parametric EWA BIC
                         "EWA_BIC" = EWA_modelling[[id]]$optim$bestval + 4*log(150),
                         "EWA_2LL" = EWA_modelling[[id]]$optim$bestval,

                         # Self-Tuning EWA (only 1 parameter)
                         #Lambda is a parameter of the softmax choice function (inverse Temperature)
                         "EWA_self_Lambda" = EWA_self_modelling[[id]]$optim$bestmem[1],
                         # Parametric EWA BIC
                         "EWA_self_BIC" = EWA_self_modelling[[id]]$optim$bestval + 1*log(150)
                         ))
}

write.csv(All_results,file="All_results.csv",row.names = FALSE)
